Hidden Talents Multiverse Analysis

### 2021-07-04 manuscript/manuscript.Rmd
final touches to manuscript

\noindent Children developing in harsh conditions often perform poorly on standard cognitive tests. However, the hidden talents approach proposes that performance for some abilities may be enhanced by adversity or improved under ecologically-relevant conditions. We examine ecologically relevant content in cognitive testing. Sampling 618 socioeconomically diverse adolescents, we measured performance on two versions of an attention shifting and working memory updating task---one abstract, the other ecological. Across multiple sources, we measured environmental unpredictability, violence, and poverty, and tested adversity × task version interactions using a preregistered multiverse approach. We did not find adversity × task version interactions for attention shifting, but there were main effects of unpredictability and violence, indicating improved attention shifting irrespective of task version. For working memory updating, violence- and poverty-exposed youth performed worse on the abstract but not the ecological version, equalizing performance among adversity-exposed youth. We found no relation between updating performance and unpredictability.

@@

# Hidden talents in context: Can ecologically-relevant stimuli improve cognitive performance among adversity-exposed youth?

@@

Although some abilities themselves are more ecologically relevant to specific dimensions of adversity, hidden talents may also emerge under ecologically relevant conditions. For example, previous studies of hidden talents have tested cognition under ecologically relevant *contexts and situations*. In particular, studies have experimentally manipulated the current context to be uncertain and mildly stressful (or not) and measured performance on executive functioning and working memory tasks[@mittal2015; @young2018] (but see[@nweze2021] for similar results without uncertainty). Strikingly, people from unpredictable environments showed improved performance on attention shifting and working memory updating tasks. However, this effect only emerged under conditions of uncertainty. There was no performance improvement under control conditions.

@@

One way to test this question is to simply replace sterile testing materials with ecologically relevant content. For example, one study tested memory and reasoning among young adults who varied in their exposures to violence[@frankenhuis2020a]. They created two versions of a memory and reasoning task, one neutral (chronological age) and one more ecologically relevant (social dominance). They found that people with more current exposures to violence were equally good, or even better, at memorizing social-dominance relationships compared to people with fewer exposures. Notably, such violence exposures were associated with worse memory for neutral (age) relationships. There were no associations revealed between violence exposures and reasoning performance. Overall, this study suggests that ecologically relevant testing materials may level the playing field for adversity-exposed youth[@ellis2020; @frankenhuis2020; @rifkin-graboi2021; @vantassel-baska2018]. In addition, although executive functions and working memory are related to many educational outcomes, it is unclear how schools and teachers might leverage skills that specifically depend on inducing uncertainty. That seems impractical, at best, and unethical, at worst.

Here we address this issue by evaluating the role of ecologically relevant content in attention shifting and working memory tasks. To do so, we sampled `r nrow(data2_non_arbitrary)` adolescents from a socioeconomically diverse population. Using interviews, self-reports, and school records, we measured exposure to environmental unpredictability, violence, and poverty. We then tested performance on two widely used tasks: attention shifting and working memory updating. There were two versions of each task. The first used standard abstract stimuli and the second replaced these stimuli with more ecologically relevant ones (see Figure \@ref(fig:figure1)). We then tested the interactive effect of task content and adversity exposure.

@@

# Results

## Overview

@@

Before conducting analyses, we identified six arbitrary data processing decisions each with two alternatives (see Arbitrary Decisions section). These criteria were 1) including the after-school club sample or not; 2) including participants with minor distractions during testing or not; 3) including participants who missed one trap question or not (those who missed more than 1 were already excluded); 4) including participants receiving special education (participants receiving more \> 60 minutes/day were already excluded) or not; 5) including participants who performed at or below chance levels of accuracy on the attention-shifting task or not (see [attention shifting data cleaning procedure]( https://osf.io/kvt5e/?view_only=92d55bd3b9834468acd3a7cef5c3febd) for more information); and 6) residualizing social desirability scores out of adversity scores or not. Using all possible combinations of arbitrary decisions, we constructed a multiverse of `r length(multi_data_list)` datasets. The minimum and maximum sample sizes were `r map_dbl(multi_data_list, "n") %>% min()` and `r map_dbl(multi_data_list, "n") %>% max()`, respectively.

@@

To facilitate interpretation of multiverse results, we report median standardized $\beta$s, median 95% confidence intervals, and percent significant for each term in our model (see Table \@ref(tab:table2)). Percent significant refers to the number of analyses (out of `r length(multi_data_list)`) that resulted in a significant effect for a particular term and reflects the relative sensitivity (or robustness) of an effect to arbitrary data processing decisions. For the main effects of adversity and task version × adversity interactions, we used a bootstrapping technique[@simonsohn2020; @orben2019; @orben2019a] to compute overall p-values, which indicate whether a $\beta$ of a term is larger than expected given the null (the median $\beta$ = 0; see Statistical Inferences).

Table \@ref(tab:table1) displays bivariate associations between all composite measures of adversity, including components of each, and descriptive statistics. Our primary interests were the main effect of adversity and the interaction of adversity with task-version. Table \@ref(tab:table2) shows median effect sizes ($\beta$-coefficients), sample sizes, and bootstrapped *p*-values for attention shifting and working memory updating analyses for the main effect of adversity (unpredictability, violence, and poverty) and the interaction term (task version × adversity). Table \@ref(tab:table3) reports median simple effects of task version across high and low levels of adversity and median simple effects of adversity across abstract and ecologically relevant task versions.

@@

We preregistered all primary analyses on the Open Science Framework (OSF; see the [osf project link](https://osf.io/6r95z/?view_only=92d55bd3b9834468acd3a7cef5c3febd), including an [updated primary analysis plan](https://osf.io/4vsnz/?view_only=92d55bd3b9834468acd3a7cef5c3febd), which included more details. We conducted secondary analyses after our primary analyses (see Secondary Analyses below). Before conducting these analyses, we uploaded secondary analysis plans (see the [first secondary analysis plan](https://osf.io/7fu35/?view_only=92d55bd3b9834468acd3a7cef5c3febd) and the [second plan](https://osf.io/wcauf/?view_only=92d55bd3b9834468acd3a7cef5c3febd), explaining the goals and purpose of each secondary analysis. A Red Team Critic[@market] successfully reproduced and verified (see this [red team report](https://osf.io/xv94z/?view_only=92d55bd3b9834468acd3a7cef5c3febd)) all analyses prior to submission.

## Primary Analyses

### Attention Shifting

Across each set of multiverse analyses, there were significant main effects of task version. All median effects sizes for task version were $\beta$s = `r c(mod_terms1$shifting.unp.type_z$median_beta, mod_terms1$shifting.vio.type_z$median_beta, mod_terms1$shifting.ses.type_z$median_beta) %>% mean()` (all *ps* \< .05) for multiverse models testing unpredictability, violence exposure, and poverty exposure, indicating that switch costs were moderately higher for the ecological version compared to the abstract version (see SI Figures 1-3). That is, reaction times on switch trials, compared to repeat trials, were longer when stimuli were ecologically relevant as opposed to abstract. However, there were no effects of age (all median $\beta$s = `r c(mod_terms1$shifting.unp.age$median_beta, mod_terms1$shifting.vio.age$median_beta, mod_terms1$shifting.ses.age$median_beta) %>% mean()`, all *ps* \>.05) or test distraction ratings (all median $\beta$ = `r c(mod_terms1$shifting.unp.test_envr$median_beta, mod_terms1$shifting.vio.test_envr$median_beta, mod_terms1$shifting.ses.test_envr$median_beta) %>% mean()`, all *ps* \> .05) on overall switch costs (see SI Figures 1-3).

@@

Overall, there were no significant interaction effects between task version and unpredictability, violence exposure, or poverty exposure (see Table \@ref(tab:table2), Table \@ref(tab:table3), and Figure \@ref(fig:figure2)b). However, analyses produced main effects of unpredictability (median $\beta$ = `r mod_terms2$shifting.unp$median_beta_main`, `r mod_terms2$shifting.unp$p_percent_main %>% str_remove("\\.00")` of *ps* \< .05, overall *p* = `r mod_terms2$shifting.unp$p_overall_main`) and violence exposure (median $\beta$ = `r mod_terms2$shifting.vio$median_beta_main`, `r mod_terms2$shifting.vio$p_percent_main` of *ps* \< .05, overall *p* = `r mod_terms2$shifting.vio$p_overall_main`). (See Table \@ref(tab:table2).) There were no main effects of poverty exposure (median $\beta$ = `r mod_terms2$shifting.ses$median_beta_main`, no *ps* \< .05, overall *p* = `r mod_terms2$shifting.ses$p_overall_main`); see Table \@ref(tab:table2)). That is, exposure to more unpredictability and violence was associated with smaller switch costs, suggesting such exposures are associated with somewhat better shifting performance on average, irrespective of whether stimuli were abstract or ecologically relevant.

@@

### Working Memory Updating

@@

Multiverse analyses revealed a main effect of poverty exposure (median $\beta$ = `r mod_terms2$updating.ses$median_beta_main`, all *ps* \< .05, bootstrapped *p* \< .001) but no main effects of unpredictability (median $\beta$ = `r mod_terms2$updating.unp$median_beta_main`, no *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.unp$p_overall_main`) or violence exposure (median $\beta$ = `r mod_terms2$updating.vio$median_beta_main`, `r mod_terms2$updating.vio$p_percent_main` of *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.vio$p_overall_main`). Analyses also revealed two significant interactions: task version × violence exposure (median $\beta$ = `r mod_terms2$updating.vio$median_beta_int`, `r mod_terms2$updating.vio$p_percent_int` of *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.vio$p_overall_int`) and task version × poverty exposure (median $\beta$ = `r mod_terms2$updating.ses$median_beta_int`, `r mod_terms2$updating.ses$p_percent_int` of *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.ses$p_overall_int`). Adolescents exposed to low levels of violence or poverty tended to perform better on the abstract working memory updating task (violence exposure median *b~abstract~* = `r simple_effects$updating.vio$beta_abs`, `r simple_effects$updating.vio$p_percent_abs` of *ps* \< .05; poverty exposure median *b~abstract~* = `r simple_effects$updating.ses$beta_abs`, all *ps* \< .05), compared to adolescents exposed to high levels of violence and poverty (see Table \@ref(tab:table3) and Figure \@ref(fig:figure3)). However, this performance gap closed on the ecological working memory task (violence exposure median *b~ecological~* = `r simple_effects$updating.vio$beta_eco`, no *ps* \< .05; poverty exposure median *b~ecological~* = `r simple_effects$updating.ses$beta_eco`, `r simple_effects$updating.ses$p_percent_eco` of *ps* \< .05). Put differently, adolescents exposed to high levels of violence or poverty ($+$ 1 *SD*) performed better on the ecologically-relevant version compared to their own performance on the abstract version (median simple *b~high~ ~violence~* = `r simple_effects$updating.vio$beta_high`, all *ps* \< .05; median simple *b~high~ ~poverty~* = `r simple_effects$updating.ses$beta_high`, all *ps* \< .05), whereas adolescents exposed to low violence or poverty ($-$ 1 *SD*) performed similarly across task versions (median simple *b~low~ ~violence~* = `r simple_effects$updating.vio$beta_low`, no *ps* \< .05; median simple *b~low~ ~poverty~* = `r simple_effects$updating.ses$beta_low`, no *ps* \< .05). In this sense, ecologically relevant stimuli appeared to equalize working memory updating performance among adolescents exposed to high levels of violence and poverty (see Table \@ref(tab:table3) and Figure \@ref(fig:figure3)a).

@@

## Secondary Analyses

We followed-up our primary analyses with two sets of secondary, exploratory analyses. We uploaded secondary analysis plans to the OSF prior to conducting them (see the [first secondary analysis plan](https://osf.io/7fu35/?view_only=92d55bd3b9834468acd3a7cef5c3febd) and the [second plan](https://osf.io/wcauf/?view_only=92d55bd3b9834468acd3a7cef5c3febd)). These analyses focused on two questions. First, in contrast to prior work[@mittal2015; @young2018], we found limited support for improved working memory updating performance in relation to unpredictability (although we found a main effect of unpredictability on attention shifting). However, prior work measured retrospective perceptions of unpredictability whereas the current combined perceptions with more objective, interview-based measures. Thus, one possibility is that specific components of our composite adversity measures (unpredictability, violence exposure, and poverty exposure) have unique associations with attention shifting and working memory updating performance. To address this, we expanded our multiverse approach to include each component of our predictor variables as additional arbitrary decisions.

@@

### Exploring Components of Adversity Dimensions

We ran an expanded multiverse analysis to explore the effects of each component of our adversity composites on attention shifting and working memory updating performance. Environmental unpredictability contained two components, perceived unpredictability (self-reported) and an interview-based measure tabulating family disruptions, parental figures, and residential moves. Violence exposure contained two self-report components: neighborhood violence and exposure to/involvement in violence. Poverty exposure contained three components: parental education/occupation (interview), perceived resource scarcity (self-report), and school-coded economic disadvantage (fee waivers, free- or reduced-priced lunch, and homelessness).

For this expanded multiverse approach, we retained all original data processing decisions (i.e., original 64 combinations of decisions) and added an adversity component as an additional decision. For unpredictability, we analyzed each unpredictability component and the composite measure (composite analyses are redundant with primary analyses but were retained for comparison purposes), which resulted in 384 analyses (64 original × three versions of unpredictability × two performance measures [attention shifting and working memory]). We applied the same approach to violence exposure (three violence variables, 384 analyses) and poverty exposure (four poverty variables, 512 analyses). The full results of these analyses are reported in the supplement (see SI Table 2).

@@

### Comparing Violence and Poverty Exposure Interactions with Task-Version

@@

Violence and poverty exposure interaction patterns were qualitatively similar to those obtained in our primary analyses (see Figure \@ref(fig:figure3)a and SI Figure 13). However, these secondary results suggest that the violence exposure interaction with task-version may be driving our poverty × task version results for working memory updating. One possibility is that the shared variance between poverty and violence exposures (*r* = `r txt5_cor$ses_vio$r`) is responsible for both interaction effects. In other words, although poverty and violence exposure are theoretically distinct, our poverty composite may capture poverty-related violence exposures (e.g., high poverty neighborhoods contain more crime) and therefore interact with task-version. When directly compared, explicit measures of violence exposure remain significant (although only under a small set of analytical decisions), whereas poverty interactions drop out entirely.

# Discussion

We examined how ecologically relevant testing materials affect cognitive performance among adversity-exposed youth. We measured three dimensions of adversity--- environmental unpredictability, violence exposure, and poverty---and tested performance on attention shifting and working memory updating tasks. We administered two versions of each task: one using abstract stimuli and the other using ecologically-relevant stimuli, allowing us to compare performance within- and between-subjects.

Leveraging multiverse analysis[@simonsohn2020; @steegen2016], we tested interactions between dimensions of adversity and task version on attention shifting and working memory updating performance. We did not find interaction effects for attention shifting. However, youth exposed to unpredictability and violence showed smaller switch costs on average, suggesting slightly improved attention shifting performance irrespective of task content. These findings partially diverge from prior work[@mittal2015; @nweze2021]. For example, whereas we found unpredictability was associated with improved attention shifting on average, Mittal and colleagues[@mittal2015] found no difference in performance under neutral conditions. And, although they improved attention shifting for those exposed to high unpredictability under uncertainty, it remains unclear how this effect relates to those obtained in the current study. In addition, Fields and colleages[@fieldsa] found caregiving instability was associated with improved attention shifting, but Nweze and colleagues[@nweze2021] found no association between adversity and attention shifting. However, they tested institutionalized children rather than measuring adversity dimensions[@nweze2021]. In general, we should interpret cautiously. Although we followed field standards for calculating switch costs by computing difference scores, such scores ignore participant-level reaction times. Mixed-modeling approaches can account for both overall reaction times and differences across switch and repeat trials, but require more statistical power to test three-way interactions. Future research should consider more appropriate modeling approaches.

For working memory updating, we found lowered performance among violence- and poverty-exposed youth on the abstract working memory updating task compared to peers with reduced exposures. However, ecological materials produced an equalization effect: the updating performance gap between low and high-adversity exposed youth narrowed on the ecological updating task. Interestingly, updating performance was unrelated to unpredictability. These findings are partially consistent with prior work on adversity exposure and working memory updating performance[@young2018; @nweze2021]. For example, both Young and colleagues[@young2018] and the current study found lowered abstract working memory performance among high adversity-exposed individuals. Although performance with ecological stimuli is not directly comparable to performance under manipulated uncertainty, both studies reduced the updating performance gap under ecologically relevant conditions. However, there are also discrepancies. For example, Young and colleagues[@young2018] found effects with unpredictability and not poverty, whereas the current study found associations with violence and poverty but not unpredictability. Finally, Nweze and colleagues[@nweze2021] found improved working memory performance among institutionalized youth without ecological stimuli or uncertainty manipulations. However, both their adversity and working memory measures were different. Such inconsistencies could be attributed to differences in sampling (adults vs. adolescents), the sampled range of adversity exposure (restricted vs. broad), or methods/design (within- vs. between-subjects design; lab vs. community/school settings). Future research should address these inconsistencies.

@@

Until recently, developmental science has almost exclusively reported lowered cognitive performance in people from harsh conditions. The hidden talents framework has challenged this notion and inspired researchers to document the strengths and abilities of people from adversity. Though much is yet to be done, our data have exciting practical implications, such as developing ecologically relevant materials for education, that might even the playing field for adversity-exposed youth. Theoretically, our data help to refine the hidden talents framework and illuminate future research questions surrounding the unique skills and abilities of people living in harsh conditions.

# Method

## Participants

@@

## Procedure

@@

## Measures

### Attention Shifting

@@

### Environmental Unpredictability

@@

### Violence Exposure.

@@

### Poverty Exposure

@@

### Control Variables

@@

## Multiverse Setup

@@

### Arbitrary Decisions.

@@

### Statistical Inferences.

@@

# References

>>>>

\noindent Children developing in harsh conditions often perform poorly on standard cognitive tests. However, the hidden talents approach proposes that performance for some abilities may be enhanced by adversity or improved under ecologically relevant conditions. We examine ecologically relevant content in cognitive testing. Sampling 618 socioeconomically diverse adolescents, we measured performance on two versions of an attention-shifting and working memory updating task---one abstract, the other ecological. Across multiple sources, we measured environmental unpredictability, violence, and poverty, and tested adversity × task version interactions using a preregistered multiverse approach. We did not find adversity × task version interactions for attention shifting, but there were main effects of unpredictability and violence, indicating improved attention shifting irrespective of task version. For working memory updating, violence- and poverty-exposed youth performed worse on the abstract but not the ecological version, equalizing performance among adversity-exposed youth. We found no relation between updating performance and unpredictability.

@@

# Hidden talents in context: Can ecologically relevant stimuli improve cognitive performance among adversity-exposed youth?

@@

Although some abilities themselves are more ecologically relevant to specific dimensions of adversity, hidden talents may also emerge under ecologically relevant conditions. For example, previous studies of hidden talents have tested cognition under ecologically relevant *contexts and situations*. In particular, studies have experimentally manipulated the current context to be uncertain and mildly stressful (or not) and measured performance on executive functioning and working memory tasks[@mittal2015; @young2018] (but see[@nweze2021] for similar results without uncertainty). Strikingly, people from unpredictable environments showed improved performance on attention-shifting and working memory updating tasks. However, this effect only emerged under conditions of uncertainty. There was no performance improvement under control conditions.

@@

One way to test this question is to simply replace sterile testing materials with ecologically relevant content. For example, one study tested memory and reasoning among young adults who varied in their exposures to violence[@frankenhuis2020a]. They created two versions of a memory and reasoning task, one neutral (chronological age) and one more ecologically relevant (social dominance). They found that people with more current exposures to violence were equally good, or even better, at memorizing social-dominance relationships compared to people with fewer exposures. Notably, such violence exposures were associated with worse memory for neutral (age) relationships. There were no associations revealed between violence exposures and reasoning performance. Overall, this study suggests that ecologically relevant testing materials may "level the playing field" for adversity-exposed youth[@ellis2020; @frankenhuis2020; @rifkin-graboi2021; @vantassel-baska2018]. In addition, although executive functions and working memory are related to many educational outcomes, it is unclear how schools and teachers might leverage skills that specifically depend on inducing uncertainty. That seems impractical, at best, and unethical, at worst.

Here we address this issue by evaluating the role of ecologically relevant content in attention-shifting and working memory tasks. To do so, we sampled `r nrow(data2_non_arbitrary)` adolescents from a socioeconomically diverse population. Using interviews, self-reports, and school records, we measured exposure to environmental unpredictability, violence, and poverty. We then tested performance on two widely used tasks: attention shifting and working memory updating. There were two versions of each task. The first used standard abstract stimuli and the second replaced these stimuli with more ecologically relevant ones (see Figure \@ref(fig:figure1)). We then tested the interactive effect of task content and adversity exposure.

@@

# Results {#results}

## Overview {#overview}

@@

Before conducting analyses, we identified six arbitrary data processing decisions each with two alternatives (see [Arbitrary Decisions](#abritrary) section). These criteria were 1) including the after-school club sample or not; 2) including participants with minor distractions during testing or not; 3) including participants who missed one trap question or not (those who missed more than 1 were already excluded); 4) including participants receiving special education (participants receiving more \> 60 minutes/day were already excluded) or not; 5) including participants who performed at or below chance levels of accuracy on the attention-shifting task or not (see [attention shifting data cleaning procedure]( https://osf.io/kvt5e/?view_only=92d55bd3b9834468acd3a7cef5c3febd) for more information); and 6) residualizing social desirability scores out of adversity scores or not. Using all possible combinations of arbitrary decisions, we constructed a multiverse of `r length(multi_data_list)` datasets. The minimum and maximum sample sizes were `r map_dbl(multi_data_list, "n") %>% min()` and `r map_dbl(multi_data_list, "n") %>% max()`, respectively.

@@

To facilitate interpretation of multiverse results, we report median standardized $\beta$s, median 95% confidence intervals, and percent significant for each term in our model (see Table \@ref(tab:table2)). Percent significant refers to the number of analyses (out of `r length(multi_data_list)`) that resulted in a significant effect for a particular term and reflects the relative sensitivity (or robustness) of an effect to arbitrary data processing decisions. For the main effects of adversity and task version × adversity interactions, we used a bootstrapping technique[@simonsohn2020; @orben2019; @orben2019a] to compute overall *p*-values, which indicate whether a $\beta$ of a term is larger than expected given the null (the median $\beta$ = 0; see [Statistical Inferences](#inference)).

Table \@ref(tab:table1) displays bivariate associations between all composite measures of adversity, including components of each, and descriptive statistics. Our primary interests were the main effect of adversity and the interaction of adversity with task-version. Table \@ref(tab:table2) shows median effect sizes ($\beta$-coefficients), sample sizes, and bootstrapped *p*-values for attention-shifting and working memory updating analyses for the main effect of adversity (unpredictability, violence, and poverty) and the interaction term (task version × adversity). Table \@ref(tab:table3) reports median simple effects of task version across high and low levels of adversity and median simple effects of adversity across abstract and ecologically relevant task versions.

@@

We preregistered all primary analyses on the Open Science Framework (OSF; see the [osf project link](https://osf.io/6r95z/?view_only=92d55bd3b9834468acd3a7cef5c3febd), including an [updated primary analysis plan](https://osf.io/4vsnz/?view_only=92d55bd3b9834468acd3a7cef5c3febd), which included more details. We conducted secondary analyses after our primary analyses (see [Secondary Analyses](#secondary) below). Before conducting these analyses, we uploaded secondary analysis plans (see the [first secondary analysis plan](https://osf.io/7fu35/?view_only=92d55bd3b9834468acd3a7cef5c3febd) and the [second plan](https://osf.io/wcauf/?view_only=92d55bd3b9834468acd3a7cef5c3febd), explaining the goals and purpose of each secondary analysis. A Red Team Critic[@market] successfully reproduced and verified (see this [red team report](https://osf.io/xv94z/?view_only=92d55bd3b9834468acd3a7cef5c3febd)) all analyses prior to submission.

## Primary Analyses {#primary}

### Attention Shifting {#primary-shifting}

Across each set of multiverse analyses, there were significant main effects of task version. All median effects sizes for task version were $\beta$s = `r c(mod_terms1$shifting.unp.type_z$median_beta, mod_terms1$shifting.vio.type_z$median_beta, mod_terms1$shifting.ses.type_z$median_beta) %>% mean()` (all *ps* \< .05) for multiverse models testing unpredictability, violence exposure, and poverty exposure, indicating that switch costs (see [Attention Shifting](#shifting) for switch cost explanation) were moderately higher for the ecological version compared to the abstract version (see SI Figures 1-3). That is, reaction times on switch trials, compared to repeat trials, were longer when stimuli were ecologically relevant as opposed to abstract. However, there were no effects of age (all median $\beta$s = `r c(mod_terms1$shifting.unp.age$median_beta, mod_terms1$shifting.vio.age$median_beta, mod_terms1$shifting.ses.age$median_beta) %>% mean()`, all *ps* \>.05) or test distraction ratings (all median $\beta$ = `r c(mod_terms1$shifting.unp.test_envr$median_beta, mod_terms1$shifting.vio.test_envr$median_beta, mod_terms1$shifting.ses.test_envr$median_beta) %>% mean()`, all *ps* \> .05) on overall switch costs (see SI Figures 1-3).

@@

Overall, there were no significant interaction effects between task version and unpredictability, violence exposure, or poverty exposure (see Table \@ref(tab:table2), Table \@ref(tab:table3), and Figure \@ref(fig:figure2)b). However, analyses produced main effects of unpredictability (median $\beta$ = `r mod_terms2$shifting.unp$median_beta_main`, `r mod_terms2$shifting.unp$p_percent_main %>% str_remove("\\.00")` of *ps* \< .05, overall *p* = `r mod_terms2$shifting.unp$p_overall_main`) and violence exposure (median $\beta$ = `r mod_terms2$shifting.vio$median_beta_main`, `r mod_terms2$shifting.vio$p_percent_main` of *ps* \< .05, overall *p* = `r mod_terms2$shifting.vio$p_overall_main`). (See Table \@ref(tab:table2).) There were no main effects of poverty exposure (median $\beta$ = `r mod_terms2$shifting.ses$median_beta_main`, no *ps* \< .05, overall *p* = `r mod_terms2$shifting.ses$p_overall_main`); see Table \@ref(tab:table2)). That is, exposure to more unpredictability and violence was associated with smaller switch costs, suggesting such exposures were associated with somewhat better shifting performance on average, irrespective of whether stimuli were abstract or ecologically relevant.

@@

### Working Memory Updating {#primary-updating}

@@

Multiverse analyses revealed a main effect of poverty exposure (median $\beta$ = `r mod_terms2$updating.ses$median_beta_main`, all *ps* \< .05, bootstrapped *p* \< .001) but no main effects of unpredictability (median $\beta$ = `r mod_terms2$updating.unp$median_beta_main`, no *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.unp$p_overall_main`) or violence exposure (median $\beta$ = `r mod_terms2$updating.vio$median_beta_main`, `r mod_terms2$updating.vio$p_percent_main` of *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.vio$p_overall_main`). Analyses also revealed two significant interactions: task version × violence exposure (median $\beta$ = `r mod_terms2$updating.vio$median_beta_int`, `r mod_terms2$updating.vio$p_percent_int` of *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.vio$p_overall_int`) and task version × poverty exposure (median $\beta$ = `r mod_terms2$updating.ses$median_beta_int`, `r mod_terms2$updating.ses$p_percent_int` of *ps* \< .05, bootstrapped *p* = `r mod_terms2$updating.ses$p_overall_int`). Adolescents exposed to low levels of violence or poverty tended to perform better on the abstract working memory updating task (violence exposure median *b~abstract~* = `r simple_effects$updating.vio$beta_abs`, `r simple_effects$updating.vio$p_percent_abs` of *ps* \< .05; poverty exposure median *b~abstract~* = `r simple_effects$updating.ses$beta_abs`, all *ps* \< .05), compared to adolescents exposed to high levels of violence and poverty (see Table \@ref(tab:table3) and Figure \@ref(fig:figure3)). However, this performance gap closed on the ecological working memory task (violence exposure median *b~ecological~* = `r simple_effects$updating.vio$beta_eco`, no *ps* \< .05; poverty exposure median *b~ecological~* = `r simple_effects$updating.ses$beta_eco`, `r simple_effects$updating.ses$p_percent_eco` of *ps* \< .05). Put differently, adolescents exposed to high levels of violence or poverty ($+$ 1 *SD*) performed better on the ecologically relevant version compared to their own performance on the abstract version (median simple *b~high~ ~violence~* = `r simple_effects$updating.vio$beta_high`, all *ps* \< .05; median simple *b~high~ ~poverty~* = `r simple_effects$updating.ses$beta_high`, all *ps* \< .05), whereas adolescents exposed to low violence or poverty ($-$ 1 *SD*) performed similarly across task versions (median simple *b~low~ ~violence~* = `r simple_effects$updating.vio$beta_low`, no *ps* \< .05; median simple *b~low~ ~poverty~* = `r simple_effects$updating.ses$beta_low`, no *ps* \< .05). In this sense, ecologically relevant stimuli appeared to equalize working memory updating performance among adolescents exposed to high levels of violence and poverty (see Table \@ref(tab:table3) and Figure \@ref(fig:figure3)a).

@@

## Secondary Analyses {#secondary}

We followed-up our primary analyses with two sets of secondary, exploratory analyses. We uploaded secondary analysis plans to the OSF prior to conducting them (see the [first secondary analysis plan](https://osf.io/7fu35/?view_only=92d55bd3b9834468acd3a7cef5c3febd) and the [second plan](https://osf.io/wcauf/?view_only=92d55bd3b9834468acd3a7cef5c3febd)). These analyses focused on two questions. First, in contrast to prior work[@mittal2015; @young2018], we found limited support for improved working memory updating performance in relation to unpredictability (although we found a main effect of unpredictability on attention shifting). However, prior work measured retrospective perceptions of unpredictability whereas the current combined perceptions with more objective, interview-based measures. Thus, one possibility is that specific components of our composite adversity measures (unpredictability, violence exposure, and poverty exposure; see [Measures](#measures) section) have unique associations with attention shifting and working memory updating performance. To address this, we expanded our multiverse approach to include each component of our predictor variables as additional arbitrary decisions.

@@

### Exploring Components of Adversity Dimensions {#secondary1}

We ran an expanded multiverse analysis to explore the effects of each component of our adversity composites (see [Measures](#measures) section) on attention shifting and working memory updating performance. Environmental unpredictability contained two components, perceived unpredictability (self-reported) and an interview-based measure tabulating family disruptions, parental figures, and residential moves. Violence exposure contained two self-report components: neighborhood violence and exposure to/involvement in violence. Poverty exposure contained three components: parental education/occupation (interview), perceived resource scarcity (self-report), and school-coded economic disadvantage (fee waivers, free- or reduced-priced lunch, and homelessness).

For this expanded multiverse approach, we retained all original data processing decisions (i.e., original 64 combinations of decisions) and added adversity component as an additional decision. For unpredictability, we analyzed each unpredictability component and the composite measure (composite analyses are redundant with primary analyses but were retained for comparison purposes), which resulted in 384 analyses (64 original × three versions of unpredictability × two performance measures [attention shifting and working memory]). We applied the same approach to violence exposure (three violence variables, 384 analyses) and poverty exposure (four poverty variables, 512 analyses). The full results of these analyses are reported in the supplement (see SI Table 2).

@@

### Comparing Violence and Poverty Exposure Interactions with Task-Version {#secondary2}

@@

Violence and poverty exposure interaction patterns were qualitatively similar to those obtained in our primary analyses (see Figure \@ref(fig:figure3)a and SI Figure 13). However, these secondary results suggest that the violence exposure interaction with task-version may be driving our poverty × task version results for working memory updating. One possibility is that the shared variance between poverty and violence exposures (*r* = `r txt5_cor$ses_vio$r`) is responsible for both interaction effects. In other words, although poverty and violence exposure are theoretically distinct, our poverty composite may capture poverty-related violence exposures (e.g., high poverty neighborhoods contain more crime) and therefore interact with task-version. When directly compared, explicit measures of violence exposure remained significant (although only under a small set of analytical decisions), whereas poverty interactions dropped out entirely.

# Discussion {#discussion}

We examined how ecologically relevant testing materials affect cognitive performance among adversity-exposed youth. We measured three dimensions of adversity--- environmental unpredictability, violence exposure, and poverty---and tested performance on attention-shifting and working memory updating tasks. We administered two versions of each task: one using abstract stimuli and the other using ecologically relevant stimuli, allowing us to compare performance within- and between-subjects.

Leveraging multiverse analysis[@simonsohn2020; @steegen2016], we tested interactions between dimensions of adversity and task version on attention shifting and working memory updating performance. We did not find interaction effects for attention shifting. However, youth exposed to unpredictability and violence showed smaller switch costs on average, suggesting slightly improved attention shifting performance irrespective of task content. These findings partially diverge from prior work[@mittal2015; @nweze2021]. For example, whereas we found unpredictability was associated with improved attention shifting on average, Mittal and colleagues[@mittal2015] found no difference in performance under neutral conditions. And, although they improved attention shifting for those exposed to high unpredictability under uncertainty, it remains unclear how this effect relates to those obtained in the current study. In addition, Fields and colleagues[@fieldsa] found caregiving instability was associated with improved attention shifting, but Nweze and colleagues[@nweze2021] found no association between adversity and attention shifting. However, they tested institutionalized children rather than measuring adversity dimensions[@nweze2021]. In general, we should interpret cautiously. Although we followed field standards for calculating switch costs by computing difference scores, such scores ignore participant-level reaction times. Mixed-modeling approaches can account for both overall reaction times and differences across switch and repeat trials, but require more statistical power to test three-way interactions. Future research should consider more appropriate modeling approaches.

For working memory updating, we found lowered performance among violence- and poverty-exposed youth on the abstract working memory updating task compared to peers with reduced exposures. However, ecological materials produced an equalization effect: the updating performance gap between low and high adversity-exposed youth narrowed on the ecological updating task. Interestingly, updating performance was unrelated to unpredictability. These findings are partially consistent with prior work on adversity exposure and working memory updating performance[@young2018; @nweze2021]. For example, both Young and colleagues[@young2018] and the current study found lowered abstract working memory performance among high adversity-exposed individuals. Although performance with ecological stimuli is not directly comparable to performance under manipulated uncertainty, both studies reduced the updating performance gap under ecologically relevant conditions. However, there are also discrepancies. For example, Young and colleagues[@young2018] found effects with unpredictability and not poverty, whereas the current study found associations with violence and poverty but not unpredictability. Finally, Nweze and colleagues[@nweze2021] found improved working memory performance among institutionalized youth without ecological stimuli or uncertainty manipulations. However, both their adversity and working memory measures were different. Such inconsistencies could be attributed to differences in sampling (adults vs. adolescents), the sampled range of adversity exposure (restricted vs. broad), or methods/design (within- vs. between-subjects design; lab vs. community/school settings). Future research should address these inconsistencies.

@@

Until recently, developmental science has almost exclusively reported lowered cognitive performance in people from harsh conditions. The hidden talents framework has challenged this notion and inspired researchers to document the strengths and abilities of people from adversity. Though much is yet to be done, our data have exciting practical implications, such as developing ecologically relevant materials for education, that might "even the playing field" for adversity-exposed youth. Theoretically, our data help to refine the hidden talents framework and illuminate future research questions surrounding the unique skills and abilities of people living in harsh conditions.

# Method {#method}

## Participants {#participants}

@@

## Procedure {#procedure}

@@

## Measures {#measures}

### Attention Shifting {#shifting}

@@

### Environmental Unpredictability {#unp}

@@

### Violence Exposure {#vio}

@@

### Poverty Exposure {#poverty}

@@

### Control Variables {#controls}

@@

## Multiverse Setup {#multiverse}

@@

### Arbitrary Decisions {#abritrary}

@@

### Statistical Inferences {#inference}

@@

# References {#refs}
